Supplementary material to the paper 'Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway

### 2023-07-31 analysis/paper/paper.Rmd
Supp2

---
title: "Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway"
# author:
#   - Isak Roalkvam, University of Oslo, Institute of Archaeology, Conservation and History
#   - Steinar Solheim, University of Oslo, Museum of Cultural History
author:
  - Isak Roalkvam:
      email: isak.roalkvam@iakh.uio.no
      institute: [iakh]
      correspondence: true
  - name: Steinar Solheim
    email: steinar.solheim@khm.uio.no
    institute: [mch]
    correspondence: false
institute:
  - iakh: University of Oslo, Institute of Archaeology, Conservation and History
  - mch: University of Oslo, Museum of Cultural History
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      fig_caption: yes
      toc: false
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - '--lua-filter=../templates/scholarly-metadata.lua'
      - '--lua-filter=../templates/author-info-blocks.lua'
      - '--lua-filter=../templates/pagebreak.lua'
header-includes:
  \usepackage{placeins}
# always_allow_html: true
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
# abstract: |
#   Text of abstract
# keywords: |
#   keyword 1; keyword 2; keyword 3
# highlights: |
#   These are the highlights. 
---

@@

# Background
Following the melting of the Fennoscandian Ice Sheet, large parts of the post-glacial landscape of Northern Scandinavia has been characterised by dramatic isostatic uplift that has led to a net sea-level fall throughout the Holocene, despite eustatic sea-level rise [e.g. @mörner1979; @steffen2011]. The majority of Mesolithic sites along the Norwegian Skagerak coast are interpreted as having been shore bound or at least located close to the contemporaneous shoreline at the time of use. As coastal foragers in this region appear to have predominantly settled on or close to the contemporaneous shoreline during early and mid-Holocene, this can be utilised to assign an approximate date to the sites. This is done by coupling the present-day altitude of the sites with reconstructions of past shoreline displacement -- a method known as shoreline dating. Shoreline dating has a long tradition of use in Norwegian archaeology [@brogger1905], and is especially useful for dating the large number of surveyed archaeological sites where temporal data that often follow with an excavation, such as radiocarbon dates or typological indicators in artefact inventories, are not available. Previous studies shave suggested that shoreline dated sites can be used as a proxy to study temporal variation in population and settlement patterns [@solheim2018; @tallavaara2020; @jorgensen2020].

Previously, the frequency of shoreline dated sites has been compared to RSPDs by finding point estimates of shoreline dates that are aggregated in somewhat arbitrary bins of 200 or 500 years. This therefore does not consider the uncertainty in the distance between the sites and the contemporaneous shoreline, nor the impact the variability in the rate of sea-level change has on the precision of the dates that can be achieved with the method. In a recent study, @roalkvam2023 has presented a probabilistic method for shoreline dating that takes these parameters into account, thus setting the stage for a more refined investigation of the relationship between frequency of ^14^C-dates and shoreline dated sites. The parametrisation of the method was based on simulating the distance between 66 sites with ^14^C-dates and the prehistoric shoreline along the Skagerrak coast in the region between Horten municipality in the northeast to Arendal municipality in the southwest (see map in Figure \@ref(fig:map)). The results of the analysis indicates that the sites tend to have been located close to the shoreline until just after 4000 BCE when a few sites become more withdrawn from the shoreline [see also @nielsen2021a], followed by a clear break around 2500 BCE, at which point shoreline dating appears to lose its utility. Thus, this geographical and temporal limit are also used for this study.   
The present paper is decidedly exploratory. While some thoughts concerning the relationship between the variables outlined above underlie the analysis, these could not be instantiated as concrete hypotheses. Rather, we explore the data by using a set of standard models to begin to unpack patterns and grasp the relationship between the variables. The dates-as-data approach is dependent on there being a link between how prehistoric people generated material that become ^14^C samples and population dynamics [cf. @freeman2018]. The sum of shoreline dated sites, on the other hand, is determined by site frequency, and its use as a population proxy is therefore dependent on there being a connection between site count and population density. If a comparison of these proxies do not, or only partially correspond, it is thus an open question what factors have impacted either distribution to cause this discrepancy, and which measure, if any, reflects true population dynamics. The issue will therefore initially demand an open and exploratory approach where a multitude of explanatory and confounding effects can be drawn on to suggest explanations of any observed pattern.

@@

For each dating method, the relative performance of the exponential, logistic and uniform models was compared. This was done by finding the overall relative log-likelihoods of the models using the *loglik* function from *ADMUR*. For any given model, this involves finding the likelihood for each shoreline date against the null-model, and in the case of the ^14^C-dates, each site-phase. The overall model likelihood is the product of each of these individual likelihoods and is not, critically [@timpson2021, 3], found by estimating the likelihood using the final SPD. From these likelihoods, the Bayesian (or Schwarz) information criterion (BIC) was then found for each model. This penalises the models for the number of parameters in use, to avoid over-fitting, and can be used to compare the relative performance of the models.

@@

All of the standard models could be rejected for the SSPD. Of the contrasted models, the BIC values indicate that the exponential model is the best candidate. The negative direction indicates that some overall process of reduction in the frequency of shoreline dated sites occurs over time, while the rejection of the model and clear deviations from the exponential patterns indicates that this model does not account for the entire development. Some significant deviations occurs with the model over-predicting the frequency of sites in the period from c. X--9000 BCE, followed by a significant positive peak just after 9000 BCE. The process of decline is then interrupted by positive deviations at, and with a smaller deviation around 7000 BCE. This is closely followed by a negative deviation of more magnitude than that expected under the exponential model between X and X BCE. Finally, a positive deviation occurs again at X--5000 BCE. While all of these deviations should be viewed with caution, the small magnitude and questions concerning the reliability of shoreline dating after 4000 BCE [cf. @roalkvam2023] means that the negative deviations that occur around 3000 BCE are disregarded here.

@@

<!-- We can also add that accepting this model means that interpreting any deviations from the Monte Carlo enevelope is inappropriate, although this has sometimes been the case. -->

@@

It is established that Stone Age sites in the region date to the period from before the earliest ^14^C-dates, and at least from as far back as around 9300 BCE [e.g. @damlien2018b; @glorstad2016]. While the SSPD achieved here starts from c. 9500 BCE, the RSPD shows no evidence of human activity until c. 8500 BCE. What has caused the lack of ^14^C-dates from this earliest period has been given many possible explanations through the years. These range from from taphonomic loss to diverse environmental and/or cultural factors, such as the use of shrubs or rapidly deteriorating seal blubber as fuel for burning, a preference for consuming raw, dried or fermented food, and possibly a lack of reliable access to firewood [see @damm2022]. While taphonomic and ecological factors cannot be excluded as part of the picture, we view it as likely that the rapid growth at the start of RSPD and the magnitude of the SSPD in the period preceding this suggest that cultural and demographic as opposed to taphonomic and environmental factors are central drivers behind the observed pattern.

The considered models indicate that after the initial jumps of each proxy, some process of decrease in shoreline dated site frequency occurs throughout the period, contrasted by a general stationary frequency of ^14^C-dated material. An immediate explanation of this mismatch could be that the SSPD is more heavily impacted by variation in land-use and mobility patterns, while the RSPD mainly reflects population numbers.

A common interpretation of the Early Mesolithic in the region is that the population was small but highly mobile, reflected in large numbers of short lived sites along the coast [e.g. @bjerck2008; @bjerck2017]. This is commonly seen in relation to homogeneous lithic technology over large parts of Scandinavia in this period, and what is believed to have been a rapid colonisation of the Norwegian coastline. High mobility has been seen as necessary both to maintain demographic viability among colonising groups, and for it to be possible to maintain this technological uniformity over large areas [@rowley-conwy2017]. The initial peak in the SSPD could thus be congruent with a process of rapid colonisation of the Norwegian coast, coupled with a high degree of mobility, leading to the overall high site-count in the earliest period. The lacking signal in the RSPD for the first settlement period following the deglaciation could in this view reflect mobility patterns characterised by short occupational duration and corresponding pyrotechnology based on small, short-lived fires [cf. @damm2022] and low population numbers.

Following this earliest phase, @berg-hansen2022 have argued that there is an increase in sites between c. 8000--7000 BCE in the northern part of the study area of this paper. However, their analysis covers a limited area and is based on the aggregation of point-based estimates of shoreline dates, which are aggregated within periods of variable length. Apart from the limitations with using point-based estimates for shoreline dates, the use of disjoint time-intervals of variable duration could cause issues for comparing the relative counts across these units [@bevan2021].

In contrast to the findings of @berg-hansen2022, the analysis undertaken here instead indicates a decline in the number of shoreline dated sites through this time-interval, with possible positive reversals of this trend just after 8000 BCE and around 7000 BCE. It has previously been suggested that a change in mobility patterns occurs from some time around 8300 BCE, involving a reduction in residential mobility [e.g. @damlien2018b; @roalkvam2022]. Furthermore, both genetic data and technological analyses of lithic inventories indicates that the initial phase of human colonisation from the south is followed by an influx, and possible mixing, of people from the east around the same time [@gunther2018; @manninen2021]. If this narrative is coupled with the data observed here, these migratory events could appear to result in a relatively sudden drop in site-count as indicated in the SSPD, which in turn could be related to a change in mobility patterns. The end of this drop also roughly corresponds to the first ^14^C-dates in the RSPD around 8300 BCE -- giving further credence to the suggestion that the appearance of material to be ^14^C-dated is mainly related to cultural and demographic processes.

The SSPD deviates negatively from all models from c. to c. BCE, and positively just before the 4th millennium BCE. The negative deviation could possibly been seen in relation to an increase in inland activity that occurs around the same time, which is argued to reach its peak around 6000 BCE. @boaz1999 related this development to a separation of inland and coastal populations in this period. The middle of the 6th millennium BCE marks the onset of the classic Nøstvet phase [@reitan2022], which has traditionally been seen as characterised by a reduction in residential mobility, and a specialisation towards fishing. This also coincides with what @fuglestvedt2018[17] has been termed \'the great wave\' of rock art in south-eastern Norway [@fuglestvedt2018, 42--46; @glorstad2010, 216--233]. Given the rejected null-models, the stable development in RSPD over this period, and the limited sample of ^14^C-dates, it is, however, difficult to

It is worth noting that there appears to be a corresponding peak in both proxies just after 8000 BCE and similarly a negative deviation just before 6000 BCE. This could possibly be related to an increase in population numbers following, and a  However, given that the logistic model could not be rejected for the RSPD, it is inappropriate to interpret any deviations from the Monte Carlo envelope, and so no analytical weight can be placed on the deviations in the RSPD given the present data. This is also poses a challenge for the interpretation of the SSPD.

Given the presence of people and corresponding signal in the SSPD before the start of the RSPD, and the suggested relevance of mobility patterns for the lacking signal in the RSPD in this early period, neither proxy is devoid of influence from population density or mobility patterns. Properly adjudicating this relationship will thus depend on independent data on mobility patterns.

@@

While population numbers would likely impact the SSPD as well, the overall mismatch between the two proxies used here means that the magnitude of demographic influence on the SSPD is not possible to ascertain.

<!-- However, the positive deviation in the SSPD from the logistic model at c. does correspond to a similar deviation of the RSPD from the exponential model. Matching negative deviations, on the other hand, can be observed around the middle of the 7th century BCE and in the period after 3500 BCE. While we are, as mentioned, cautions in giving these much weight in the SSPD alone, the correspondence of these deviations across the two proxies gives some reason to suggests that these might reflect meaningful demographic signals. This would also i -->

While it appears that the SSPD and RSPD are differentially impacted by population density, and, we cautiously suggest, the SSPD is more heavily influenced by land-use and mobility patterns, this suggestion and the nature of the discrepancy between the proxies clearly requires further investigation. Some lines of inquiry could be to also draw on lithic technology and tool-kit diversity as a function of population size and density [e.g. @collard2013a; @riede2014; @solheim2020], and mobility patterns that can potentially be inferred from the composition of lithic assemblages [e.g. @barton2014; @clark2017; @preston2018; @roalkvam2022]. Environmental conditions, constraints and developments also represent a range of factors that can have implications for these dimensions, and will consequently be of future importance to unpack the relationship between the proxies [e.g. @jorgensen2020; @ordonez2022]. Aligning these different lines of evidence would also be dependent on a more direct modelling of the development in the SSPS, rather than simply basing the analysis on deviations from a rejected null model [cf. @timpson2021]. Furthermore, mapping and contrasting these developments across a larger region will not only offer a larger sample size, which is especially limited for the ^14^C-data, but could also provide important contextualisation for the developments observed here [see @lundstrom2021].

@@

# Conclusion
The finding that the frequency of radiocarbon dates and frequency of shoreline dated sites, in general, follow inverse trajectories is a valuable insight. Although a more precise understanding of this relationship

>>>>

---
title: "Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway"
# author:
#   - Isak Roalkvam, University of Oslo, Institute of Archaeology, Conservation and History
#   - Steinar Solheim, University of Oslo, Museum of Cultural History
author:
  - Isak Roalkvam:
      email: isak.roalkvam@iakh.uio.no
      institute: [iakh]
      correspondence: true
  - name: Steinar Solheim
    email: steinar.solheim@khm.uio.no
    institute: [mch]
    correspondence: false
institute:
  - iakh: University of Oslo, Institute of Archaeology, Conservation and History
  - mch: University of Oslo, Museum of Cultural History
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      fig_caption: yes
      toc: false
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - '--lua-filter=../templates/scholarly-metadata.lua'
      - '--lua-filter=../templates/author-info-blocks.lua'
      - '--lua-filter=../templates/pagebreak.lua'
header-includes:
  \usepackage{placeins}
  \usepackage[left]{lineno}
  \linenumbers
# always_allow_html: true
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
# abstract: |
#   Text of abstract
# keywords: |
#   keyword 1; keyword 2; keyword 3
# highlights: |
#   These are the highlights. 
---

@@

# Background
Following the melting of the Fennoscandian Ice Sheet, large parts of the post-glacial landscape of Northern Scandinavia has been characterised by dramatic isostatic uplift that has led to a net sea-level fall throughout the Holocene, despite eustatic sea-level rise [e.g. @mörner1979; @steffen2011]. The majority of Mesolithic sites along the Norwegian Skagerak coast are interpreted as having been shore bound or at least located close to the contemporaneous shoreline at the time of use. As coastal foragers in this region appear to have predominantly settled on or close to the contemporaneous shoreline during early and mid-Holocene, this can be utilised to assign an approximate date to the sites. This is done by coupling the present-day altitude of the sites with reconstructions of past shoreline displacement -- a method known as shoreline dating. Shoreline dating has a long tradition of use in Norwegian archaeology [@brogger1905], and is especially useful for dating the large number of surveyed archaeological sites where temporal data that often follow with an excavation, such as radiocarbon dates or typological indicators in artefact inventories, are not available. Previous studies have suggested that shoreline dated sites can be used as a proxy to study temporal variation in population and settlement patterns [@jorgensen2020; @mjærum2022; @solheim2018; @tallavaara2020].

Previously, the frequency of shoreline dated sites has been compared to RSPDs by finding point estimates of shoreline dates that are aggregated in somewhat arbitrary bins of 200 or 500 years. This therefore does not consider the uncertainty in the distance between the sites and the contemporaneous shoreline, nor the impact the variability in the rate of sea-level change has on the precision of the dates that can be achieved with the method. In a recent study, @roalkvam2023 has presented a probabilistic method for shoreline dating that takes these parameters into account, thus setting the stage for a more refined investigation of the relationship between frequency of ^14^C-dates and shoreline dated sites. The parametrisation of the method was based on simulating the distance between 66 sites with ^14^C-dates and the prehistoric shoreline along the Skagerrak coast in the region between Horten municipality in the northeast to Arendal municipality in the southwest (see map in Figure \@ref(fig:map)). The results of the analysis indicates that the sites tend to have been located close to the shoreline until just after 4000 BCE when a few sites become more withdrawn from the shoreline [see also @nielsen2021a], followed by a clear break around 2500 BCE, at which point shoreline dating appears to lose its utility. Thus, this geographical and temporal limit are also used for this study.

The present study is decidedly exploratory. While some thoughts concerning the relationship between the variables outlined above underlie the analysis, these could not be instantiated as concrete hypotheses. Rather, we explore the data by using a set of standard models to begin to unpack patterns and grasp the relationship between the variables. The dates-as-data approach is dependent on there being a link between how prehistoric people generated material that become ^14^C samples and population dynamics [cf. @freeman2018]. The sum of shoreline dated sites, on the other hand, is determined by site frequency, and its use as a population proxy is therefore dependent on there being a connection between site count and population density. If a comparison of these proxies do not, or only partially correspond, it is thus an open question what factors have impacted either distribution to cause this discrepancy, and which measure, if any, reflects true population dynamics. The issue will therefore initially demand an open and exploratory approach where a multitude of explanatory and confounding effects can be drawn on to suggest explanations of any observed pattern.

@@

For each dating method, the relative performance of the exponential, logistic and uniform models was compared. This was done by finding the overall relative log-likelihoods of the models using the *loglik* function from *ADMUR*. For any given model, this involves finding the scalar product for each shoreline date against the probability of the null-model, and in the case of the ^14^C-dates, each site-phase. The overall model likelihood is the product of these individual probability and is not, critically [@timpson2021, 3], found by using the final SPD. From these likelihoods, the Bayesian (or Schwarz) information criterion (BIC) was then found for each model. This penalises the models for the number of parameters in use, to avoid over-fitting, and can be used to compare the relative performance of the models.

@@

All of the standard models could be rejected for the SSPD. Of the contrasted models, the BIC values indicate that the exponential model is the best candidate. Its negative direction indicates that some overall process of reduction in the frequency of shoreline dated sites occurs over time, while the rejection of the model and clear deviations from the exponential patterns indicates that this model does not account for the entire development. Some significant deviations occurs with the model over-predicting the frequency of sites in the period from c. X--9000 BCE, followed by a significant positive peak just after 9000 BCE. The process of decline is then interrupted by positive deviations at, and with a smaller deviation around 7000 BCE. This is closely followed by a negative deviation of more magnitude than that expected under the exponential model between X and X BCE. Finally, a positive deviation occurs again at X--5000 BCE. While all of the above deviations should also be considered with caution, questions concerning the reliability of shoreline dating after 4000 BCE [cf. @roalkvam2023] and the small magnitude of the negative deviations that occur around 3000 BCE means that are disregarded here.

@@

<!-- We can also add that accepting this model means that interpreting any deviations from the Monte Carlo enevelope is inappropriate, although this has sometimes been the case. -->
```{r bic, echo = FALSE, fig.cap = "Bayesian information criterion for the standard models fit to the SPDs.", out.width = "500px", fig.align="center", regfloat = TRUE}
knitr::include_graphics(here("analysis/figures/bic.png"))
```

@@

It is established that Stone Age sites in the region date to the period from before the earliest ^14^C-dates, and at least from as far back as around 9300 BCE [e.g. @damlien2018b; @glorstad2016]. While the SSPD achieved here starts from c. 9500 BCE, the RSPD shows no evidence of human activity until c. 8600 BCE. What has caused the lack of ^14^C-dates from this earliest period has been given many possible explanations through the years. These range from taphonomic loss to diverse environmental and/or cultural factors, such as the use of shrubs or rapidly deteriorating seal blubber as fuel for burning, a preference for consuming raw, dried or fermented food, and possibly a lack of reliable access to firewood [see @damm2022]. While taphonomic and ecological factors cannot be excluded as part of the picture, we view it as likely that the rapid growth at the start of RSPD and the magnitude of the SSPD in the period preceding this suggest that cultural and demographic factors are central drivers behind the observed pattern.

The considered models also indicate that after the initial jumps of each proxy, some process of decrease in shoreline dated site frequency occurs throughout the period, contrasted by a general stationary frequency of ^14^C-dated material. An immediate explanation of this mismatch could be that the SSPD is more heavily impacted by variation in land-use and mobility patterns, while the RSPD mainly reflects population numbers.

A common interpretation of the Early Mesolithic in the region is that the population was small but highly mobile, reflected in large numbers of short-lived sites along the coast [e.g. @bjerck2008; @bjerck2017]. This is commonly seen in relation to homogeneous lithic technology over large swathes of the Scandinavian Peninsula in this period, and what is believed to have been a rapid colonisation of the Norwegian coastline [e.g. @bang-andersen2012; @bjerck2009; @fuglestvedt2012]. High mobility has been seen as necessary both to maintain demographic viability among colonising groups, and for it to be possible to maintain this technological uniformity over large areas [@rowley-conwy2017; also discussed by @berg-hansen2018]. The initial peak in the SSPD could thus be congruent with a process of rapid colonisation of the Norwegian coast, coupled with a high degree of mobility, leading to the overall high site-count in the earliest period. The lacking signal in the RSPD for the first settlement period following the deglaciation could in this view reflect mobility patterns characterised by short occupational duration and corresponding pyrotechnology based on small, short-lived fires [cf. @damm2022] and low population numbers.

Following this earliest phase of rapid increase in shoreline dated sites, there a drop from this initial peak starts around 8800 BCE and concludes around 8000 BCE. It has previously been suggested that a change in mobility patterns occurs from some time around 8300 BCE, involving a reduction in residential mobility [e.g. @damlien2018b; @roalkvam2022]. Furthermore, both genetic data and technological analyses of lithic inventories indicates that the initial phase of human colonisation from the south is followed by an influx, and possible mixing, of people from the east around the same time [@günther2018; @manninen2021]. If this narrative is coupled with the data observed here, these migratory events could appear to result in a relatively sudden drop in site-count as indicated in the SSPD, which in turn could be related to a change in mobility patterns. The end of this drop also roughly corresponds to the first ^14^C-dates in the RSPD around 8300 BCE, leading to the logistic growth that plateaus in the first centuries after 8000 BCE. This logistic growth in the RSPD has a corresponding positive deviation in the SSPD, which could indicate that this demographic signal is reflected in the frequency of shoreline dated sites as well.

@berg-hansen2022 argued that there is an increase in sites between c. 8000--7000 BCE in the northern part of the study area of this paper. Their analysis covers a more limited area and is based on the aggregation of point-based estimates of shoreline dates, which are aggregated within periods of variable length. Apart from the limitations with using point-based estimates for shoreline dates, the use of disjoint discrete time-intervals of variable duration could cause issues for comparing the relative counts across these units [@bevan2021]. However, in line with the findings of @berg-hansen2022, the analysis undertaken here indicates a positive and quite marked reversal of the declining trend both around 8000--7500 BCE and with a smaller magnitude around 7100--6900 BCE.

Subsequently, the SSPD deviates negatively from the exponential decay from c. 6800 to c. 5800 BCE, a decrease that was also noted by @berghansen2022. This is followed by a positive deviation just before the 4th millennium BCE. The negative deviation could possibly been seen in relation to an increase in inland activity in southern Norway that occurs around the same time, which is argued to reach its peak around 6000 BCE. It is also worth noting that there appears to be a corresponding negative deviation from the logistic null model in the RSPD just before 6000 BCE. However, given that the model could not be rejected for the RSPD, it is inappropriate to interpret any deviations from the Monte Carlo envelope, and so no analytical weight can be placed on the deviations in the RSPD, given the present data.

@boaz1999 related the development in inland activity to a separation of inland and coastal populations in this period. This is followed by the onset of the classic Nøstvet phase in the middle of the 6th millennium BCE  [@reitan2022], which has traditionally been seen as characterised by a reduction in residential mobility, and a specialisation towards fishing. This also coincides with what @fuglestvedt2018[17] has been termed \'the great wave\' of rock art in south-eastern Norway [@fuglestvedt2018, 42--46; @glorstad2010, 216--233]. The developments that might be reflected in the positive deviation in the SSPD.

Given the rejected null-models, the stable development in RSPD over this period, and the limited sample of ^14^C-dates, it is, however, difficult to

Given the presence of people and corresponding signal in the SSPD before the start of the RSPD, and the suggested relevance of mobility patterns for the lacking signal in the RSPD in this early period, neither proxy is devoid of influence from population density or mobility patterns. Furthermore, the correspondence between the logistic growth in the RSPD and the positive deviation in the SSPD could suggest that after However, this has to be left as very tenuous suggestions, as properly adjudicating this relationship will depend on independent data that tracks the developments of the other factors that hvae been suggested to be of relevance to the behaviour of the two proxies.

@@

While population numbers would likely impact the SSPD as well, the overall mismatch between the two proxies used here means that the magnitude of demographic influence on the SSPD is not possible to ascertain. 
<!-- However, the positive deviation in the SSPD from the logistic model at c. does correspond to a similar deviation of the RSPD from the exponential model. Matching negative deviations, on the other hand, can be observed around the middle of the 7th century BCE and in the period after 3500 BCE. While we are, as mentioned, cautions in giving these much weight in the SSPD alone, the correspondence of these deviations across the two proxies gives some reason to suggests that these might reflect meaningful demographic signals. This would also i -->

Thus, while it appears that the SSPD and RSPD are differentially impacted by population density, and, we cautiously suggest, the SSPD is more heavily influenced by land-use and mobility patterns, this suggestion and the nature of the discrepancy between the proxies clearly requires further investigation. Some lines of inquiry could be to also draw on lithic technology and tool-kit diversity as a function of population size and density [e.g. @berg-hansen2018; @collard2013a; @riede2014; @solheim2020], and mobility patterns that can potentially be inferred from the composition of lithic assemblages [e.g. @barton2014; @clark2017; @preston2018; @roalkvam2022]. Environmental conditions, constraints and developments also represent a range of factors that can have implications for these dimensions, and will consequently be of future importance to unpack the relationship between the proxies [e.g. @jorgensen2020; @ordonez2022]. Aligning these different lines of evidence would also be dependent on a more direct modelling of the development in the SSPD, rather than simply basing the analysis on deviations from a rejected null model [cf. @timpson2021]. Furthermore, mapping and contrasting these developments across a larger region will not only offer a larger sample size, which is especially limited for the ^14^C-data, but could also provide important contextualisation for the developments observed here [see @lundstrom2021].

@@

# Conclusion

This paper has presented a first exploration of summing the results of using the probabilistic method for shoreline dating presented in @roalkvam2023a for assessing the development of the frequency of shoreline dated sites over time. The approach demonstrates that it

The finding that the frequency of radiocarbon dates and frequency of shoreline dated sites, in general, follow trajectories is a valuable insight. Although a more precise understanding of this relationship

### 2023-07-31 analysis/supplementary/supplementary.Rmd
Supp2

```{r data, echo = FALSE}
# For the compilation of this Rmarkdown document, the results were pre-compiled 
# and are loaded here. Set eval = TRUE in code chunks below to re-run the code.
load(file = here::here("analysis/data/derived_data/shore_pd_models.RData"))

>>>>

```{r precompiled, echo = FALSE}
# For the compilation of this Rmarkdown document, the results were pre-compiled 
# and are loaded here. Set eval = TRUE in code chunks below to re-run the code.
load(file = here::here("analysis/data/derived_data/supplementary.RData"))
